Generalized Plackett-Luce Likelihoods

### 2022-08-04 inst/jss_hyper3.Rnw
further work on JSS manuscript

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do
eiusmod tempor incididunt ut labore et dolore magna aliqua.  Ut enim
ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut
aliquip ex ea commodo consequat.  Duis aute irure dolor in
reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla
pariatur.  Excepteur sint occaecat cupidatat non proident, sunt in
culpa qui officia deserunt mollit anim id est laborum

The \code{hyper2} package~\citep{hankin2009,hankin2017} furnishes
computational support for generalized
Plackett-Luce~\citep{plackett1975} likelihood functions.  Our
preferred interpretation is a race (as in track and field atheletics):
given six competitors $1-6$, we ascribe to them nonnegative strengths
$p_1\ldots p_6$; the probability that $i$ beats $j$ is
$p_i/(p_i+p_j)$.  It is conventional to normalise so that the total
strength is unity, and to identify a competitor with his strength.
Given an order statistic, say $p_1\succ p_2\succ p_3\succ p_4$, the
Plackett-Luce likelihood function would be

\begin{equation}
  \frac{p_1}{p_1+p_2+p_3+p_4}\cdot
  \frac{p_2}{    p_2+p_3+p_4}\cdot
  \frac{p_3}{        p_3+p_4}\cdot
  \frac{p_4}{            p_4}.
\end{equation}

@@

However, consider a race in which there are six runners 1-6 but we
happen to know that three of the runners (1,2,3) are clones of
strength $p_a$, say, two of the runners (4,5) have strength $p_b$, and
the final runner (6) is of strength $p_c$.  The runners race and the
finishing order is:

@@

\begin{equation}
\frac{p_a}{3p_a+2p_b+p_c}\cdot
\frac{p_c}{2p_a+2p_b+p_c}\cdot
\frac{p_b}{2p_a+2p_b    }\cdot
\frac{p_a}{2p_a+ p_b    }\cdot
\frac{p_a}{ p_a+ p_b    }\cdot
\frac{p_b}{      p_b    },\qquad p_a+p_b+p_c=1
\end{equation}

Existing \code{hyper2} formalism as described by \citet{hankin2017}
cannot represent this likelihood function, because
Equation~\ref{hyper2likelihood} uses sets as the indexing elements,
and in this case we need multisets\footnote{Note that the version of
\code{hyper2} presented by \citet{hankin2017} used integer-valued sets
together with a print method that used a complicated mapping system
from integers to competitor names.  Current methodology [following
commit \code{51a8b46}] is to use sets of character strings which
represent the competitors directly; this allows for easier combination
of observations including different competitors.}.

<<>>=
2+2
@

>>>>

<<setup, cache=FALSE,echo=FALSE>>=
library("knitr")
opts_chunk$set(cache=TRUE, autodep=TRUE)
@

\section{introduction}

The \code{hyper2} package~\citep{hankin2009,hankin2017} furnishes
computational support for generalized
Plackett-Luce~\citep{plackett1975} likelihood functions.  Our
preferred interpretation is a race (as in track and field atheletics):
given six competitors $1-6$, we ascribe to them nonnegative strengths
$p_1,\ldots, p_6$; the probability that $i$ beats $j$ is
$p_i/(p_i+p_j)$.  It is conventional to normalise so that the total
strength is unity, and to identify a competitor with his strength.
Given an order statistic, say $p_1\succ p_2\succ p_3\succ p_4$, the
Plackett-Luce likelihood function would be

\begin{equation}\label{PL_like}
  \frac{p_1}{p_1+p_2+p_3+p_4}\cdot
  \frac{p_2}{    p_2+p_3+p_4}\cdot
  \frac{p_3}{        p_3+p_4}\cdot
  \frac{p_4}{            p_4}.
\end{equation}

@@

However, consider a race in which there are six runners 1-6 but we
happen to know that three of the runners (1,2,3) are clones of
strength $p_a$, two of the runners (4,5) have strength $p_b$, and the
final runner (6) is of strength $p_c$.  We normalise so
$p_a+p_b+p_c=1$.  The runners race and the finishing order is:

@@

\begin{equation}\label{plackettluce}
\frac{p_a}{3p_a+2p_b+p_c}\cdot
\frac{p_c}{2p_a+2p_b+p_c}\cdot
\frac{p_b}{2p_a+2p_b    }\cdot
\frac{p_a}{2p_a+ p_b    }\cdot
\frac{p_a}{ p_a+ p_b    }\cdot
\frac{p_b}{      p_b    },\qquad p_a+p_b+p_c=1
\end{equation}

Existing \code{hyper2} formalism as described by \citet{hankin2017}
cannot represent this likelihood function, because
Equation~\ref{hyper2likelihood} uses sets as the indexing elements,
and in this case we need multisets\footnote{Note that the version of
\code{hyper2} presented by \citet{hankin2017} used integer-valued sets
together with a print method that used a complicated mapping system
from integers to competitor names.  Current methodology [following
commit \code{51a8b46}] is to use sets of character strings which
represent the competitors directly; this allows for easier combination
of observations including different competitors.}.  The declarations

\begin{verbatim}
typedef map<string, long double> weightedplayervector;
typedef map<weightedplayervector, long double> hyper3;
\end{verbatim}

show how the \proglang{map} class of the Standard Template Library are
used: \proglang{weightedplayervector} objects map strings to long
doubles (specifically, mapping player names to their multiplicities),
and objects of class \proglang{hyper3} are maps from a
\proglang{weightedplayervector} object to long doubles.  One advantage
of this is efficiency: Search, removal, and insertion operations have
logarithmic complexity.  As an example, the following \proglang{C++}
pseudo code would create a log-likelihood function for the first term
in Equation~\ref{plackettluce}:

\begin{verbatim}
weightedplayervector n,d;
n["a"] = 1; // n for numerator
d["a"] = 3; // d for denominator
d["b"] = 2;
d["c"] = 1;

hyper3 L;
L[n] = 1;
L[d] = -1;
\end{verbatim}

Above, object \code{L} is an object of class \code{hyper3}; it may be
evaluated at points in probability space [that is, a vector
\code{[a,b,c]} of nonnegative values with unit sum] using standard
\proglang{R} idiom wrapping \proglang{C++} back end.

The package includes an \proglang{S4} class \code{hyper3} for this
type of object.  Package idiom for creating such objects uses named
vectors:

<<>>=
library("hyper2")
H <- hyper3()
H[c(a=1)] <- 1
H[c(a=3,b=2,c=1)] <- -1
H
@

Above, we see object \code{H} is a log-likelihood function of the
players' strengths, which may be evaluated at specified points in
probability space.  If, for example, we wish to assess $H_1\colon
p_a=0.9,p_b=0.05,p_c=0.05$ and $H_2\colon p_a=0.01,p_b=0.01,p_c=0.98$,
we can evaluate these hypotheses using generic function
\code{loglik()}:

<<>>=
loglik(c(a=0.01,b=0.01,c=0.98),H)
loglik(c(a=0.9,b=0.05,c=0.05),H)
@

Thus we prefer $H_1$ over $H_2$ with about 3.5 units of support,
satisfying the standard two units of support
criterion~\citep{edwards1972}, and we conclude that our observation
[in this case, that one of the three clones of player $a$ beat the $b$
twins and the singleton $c$] furnishes strong support against $H_2$ in
favour of $H_1$.

The package includes many helper functions to work with order
statistics of this type.  Function \code{ordervec2supp3()}, for
example, can be used to generate a loglikelihood function for
Equation~\ref{PL_like}:

<<>>=
(H <- ordervec2supp3(c("a","c","b","a","a","b")))
@

We may find a maximum likelihood estimate for the players' strengths:

<<maxexamp,cache=TRUE>>=
(mH <- maxp(H))
@

We may use a selection of statistical tests on the order statistic,
modified from~\citep{hankin2017} to work with \code{hyper3} objects.
For example, we may assess the hypothesis that all three players are
of equal strength [viz $H_0\colon p_a=p_b=p_c=\frac{1}{3}$]:

<<testequality,cache=TRUE>>=
equalp.test(H)
@

showing, perhaps unsurprisingly, that this small dataset is consistent
with $H_0$.

\section{Some simple cases}

Here I consider some order statistics with nontrivial maximum
likelihood Bradley-Terry strength.  The simplest nontrivial case would
be three competitors with strengths $a,a,b$ and finishing order
$a\succ b\succ a$.  The likelihood function would be

\begin{equation}\label{aba}
\frac{a}{2a+b}\cdot\frac{b}{a+b}
\end{equation}

In this case we know that $a+b=1$ so this is equal to
$\mathcal{L}=\mathcal{L}(a)=\frac{a(1-a)}{1+a}$.

\begin{equation}\label{aba_mle}
\frac{d\mathcal{L}}{da}=\frac{(1+a)(1-2a)-a(1-a)}{(1+a)^2}=
\frac{1-2a+a-2a^2 -a + a^2}{(1+a)^2}=
\frac{1-2a-a^2}{(1+a)^2}
\end{equation}

This will be zero at $\sqrt{2}\pm 1$; we also note that
$d^2\mathcal{L}/da^2=-4(1+a)^{-3}$, manifestly strictly negative for
$0\leq a\leq 1$: the root is a maximum.

<<maxpaba,cache=TRUE>>=
maxp(ordervec2supp3(c("a","b","a")))
@

above, we see close agreement with the theoretical value of
$(\sqrt{2}-1,1-\sqrt{2})\simeq (0.414,0.586)$.  Observe that the
maximum likelihood estimate for $a$ is strictly less than 0.5, even
though the finishing order is symmetric.  Using
$\mathcal{L}(a)=\frac{a(1-a)}{1+a}$, we can show that
$\log\mathcal{L}(\hat{a})=\log\left(3-2\sqrt{2}\right)\simeq -1.76$,
where $\hat{a}=\sqrt{2}-1$ is the maximum likelihood estimate for $a$.
Defining $\mathcal{S}=\log\mathcal{L}$ as the support [log-likelihood]
we have

\begin{equation}
\mathcal{S}=\mathcal{S}(a)=\log\left(\frac{a(1-a)}{1+a}\right)-\log\left(3-2\sqrt{2}\right)
\end{equation}

as a standard support function which has a maximum value of zero when
evaluated at $\hat{a}=\sqrt{2}-1$.  For example, we can test the null
that $a=b=\frac{1}{2}$, the statement that the competitors have equal
Bradley-Terry strengths:

<<>>=
a <- 1/2 # null
(S_delta <- log( a*(1-a)/(1+a))- log(3-2*sqrt(2)))
@

Thus the additional support gained in moving from $a=\frac{1}{2}$ to
the evaluate of $a=\sqrt{2}-1$ is 0.029, rather small [as might be
expected given that we have only one rather uninformative observation,
and also given that the maximum likelihood estimate ($\simeq 0.41$) is
quite close to the null of $0.5$].  Nevertheless we can apply Wilks's
theorem for a $p$-value:

<<>>=
pchisq(-2*S_delta,df=1,lower.tail=FALSE)
@

The $p$-value is about 0.81, exceeding 0.05; thus we have no strong
evidence to reject the null of $a=\frac{1}{2}$.  The observation is
informative, in the sense that we can find a credible interval for
$a$.  With an $n$-unit of support criterion the analytical solution to
$\mathcal{S}(p)=-n$ is given by defining $X=\log(3-2\sqrt{2})-n$ and
solving $p(1-p)/(1+p)=X$, or
$p_\pm=\left(1-X\pm\sqrt{1+4X+X^2}\right)/2$, the two roots being the
lower and upper limits of the credible interval;
Figure~\ref{ABA_likelihood}.

\begin{figure}[htbp]
  \begin{center}
<<figaba>>=
a <- seq(from=0,by=0.005,to=1)
S <- function(a){log(a*(1-a)/((1+a)*(3-2*sqrt(2))))}
plot(a,S(a),type='b')
abline(h=c(0,-2))
abline(v=c(0.02438102,0.9524271),col='red')
abline(v=sqrt(2)-1)
@
\caption{A\label{ABA_likelihood} support function for $p_a$ with observation $a\succ b\succ a$}
\end{center}
\end{figure}

\subsection{All results for AAB}

\begin{itemize}
\item $A\succ A\succ B\longrightarrow\frac{a^2}{1+a}$
\item $A\succ B\succ A\longrightarrow\frac{1-2a-a^2}{(1+a)^2}$
\item $B\succ A\succ A\longrightarrow\frac{1-a}{1+a}$
\end{itemize}

Likelihood functions for these order statistics are given in
Figure~\ref{ABAetc_likelihood}.  It can be shown that the Fisher
information for such observations is

\begin{equation}
\mathcal{I}(a)=\frac{1+16a+18a^2-4a^3-15a^4-7a^5-a^6}{a(1-a)(1+a)^3}
\end{equation}

which has a minimum at about $0.575$.

\begin{figure}[htbp]
  \begin{center}
<<figabbabb,echo=FALSE>>=
f_aab <- function(a){a^2/(1+a)}
f_aba <- function(a){a*(1-a)/(1+a)^2}
f_baa <- function(a){(1-a)/(1+a)}

p <- function(f,...){points(a,f(a)/max(f(a)),...)}
plot(0:1,0:1,xlab="p(a)",ylab="Likelihood",type="n")
p(f_aab,type="l",col="black")
p(f_aba,type="l",col="red")
p(f_baa,type="l",col="blue")
legend(x=0.85,y=0.7,col=c("black","red","blue"),lty=1,pch=NA,legend=c("AAB","ABA","BAA"))
abline(h=exp(-2),lty=2)
@
\caption{Likelihood \label{ABAetc_likelihood} functions for
observations $a\succ a\succ b$, $a\succ b\succ a$, $b\succ a\succ a$}
\end{center}
\end{figure}

\subsection{Nonfinishers}

If we allow non-finishers [that is, a subset of competitors who are
classified as being beaten by any finishing competitors], there is
another nontrivial order statistic, viz $a\succ b\succ\left\lbrace
a,b\right\rbrace$ [thus one of the two $a$'s won, one of the $b$'s
came second, and one of each of $a$ and $b$ failed to finish].

\begin{equation}
\mathcal{L}(a)=
\frac{a}{2a+2b}\cdot
\frac{b}{ a+2b}\propto\frac{a(1-a)}{2-a}
\end{equation}

(see how the likelihood function is actually simpler than for the
complete order statistic).  The evaluate would be $2-\sqrt{2}\simeq
0.586$:

<<cache=TRUE>>=
maxp(ordervec2supp3(c("a","b"),nonfinishers=c("a","b")))
@

\section{Mann-Whitney test}

The ideas presented above can easily be extended to arbitrarily large
numbers of competitors.  Given an order statistic of the type
considered above, the Mann-Whitney-Wilcoxon test assesses a null of
identity of underlying distributions.  Consider the fly spray dataset
presented by \cite{wilcoxon1945}:

<<defineAB>>=
A <- c(68,68,59,72,64,67,70,74)
B <- c(60,67,61,62,67,63,56,58)
@

Above we see the percentage mortality of two preparations, A and B,
each applied eight times.  We may easily convert this dataset to a
rank order statistic:

<<makeABorderstatistic>>=
names(A) <- rep("a",length(A))
names(B) <- rep("b",length(B))
sort(c(A,B))
names(sort(c(A,B)))
@

And thence to a likelihood function:

<<definehyper3AB,cache=TRUE>>=
(Hab <- ordervec2supp3(names(sort(c(A,B)))))
@

Package idiom for testing the null of equality is straightforward:

<<testABforequality,cache=TRUE>>=
equalp.test(Hab)
@

Compare \citeauthor{wilcoxon1945}'s $p$-value of $1.77\%$.  However,
observe that the \code{hyper3} likelihood approach gives more
information than Wilcoxon's analysis: Firstly, we see from the
evaluate that the probability of a single application of preparation A
killing more flies than B is about 15\%.  Further, we may plot a
support curve for this dataset as above:

\begin{figure}[htbp]
  \begin{center}
<<plotwilcoxlike,cache=TRUE>>=
a <- seq(from=0.02,to=0.5,len=50)
L <- sapply(a,function(p){loglik(p,Hab)})
plot(a,L-max(L),type='b')
abline(h=c(0,-2))
abline(v=0.1558)
@
\caption{A\label{wilcox_likelihood} support for the Bradley-Terry
strength $p_a$ of preparation A~\citep{wilcoxon1945}.  The evaluate of
0.1558 is shown; and the two-units-of support credible interval, which
excludes $p_a=0.5$}
\end{center}
\end{figure}

\section{Multiple competitors: javelin}

The ideas presented above are easily extended to multiple types of
competitors.  Consider the following table, drawn from the men's
javelin, 2020 Olympics:

<<javelintable>>=
javelin_table
@

This Chopra threw 87.03m on his first throw, 87.58m on his second, and
so on.  No-throws, ignored here, are indicated with a \code{X}.  We
may convert this to a named vector with elements being the throw
distances, and names being the competitors, using
`attemptstable2supp3()`:

<<label=converttosupp3,cache=TRUE>>=
javelin_vector <- attemptstable2supp3(javelin_table,decreasing=TRUE,give.supp=FALSE)
javelin_vector
@

Above we see that Chopra threw the longest and second-longest throws
of 87.58m and 87.03 respectively; Vadlejch threw the third-longest
throw of 86.67m, and so on (\code{NA} entries correspond to
no-throws.)  Now convert the attempts table to a \code{hyper3} object,
again using function \code{attemptstable2supp3()} but this time pass
\code{give.supp=TRUE}:

<<dothething2,cache=TRUE>>=
javelin <- ordervec2supp3(v=names(javelin_vector)[!is.na(javelin_vector)])
@

Above, object \code{javelin} is a \code{hyper3} likelihood function,
so one has access to the standard likelihood-based methods, such as
finding and displaying the maximum likelihood estimate:

<<setdigits,echo=FALSE>>=
options(digits=3)
@

\begin{figure}[htbp]
  \begin{center}
<<testthejav,cache=TRUE>>=
(mj <- maxp(javelin))
dotchart(mj,pch=16)
@
\caption{Maximum likelihood estimate for javelin throwers' Bradley-Terry strengths}
\end{center}
\end{figure}

Above, we see that Vadlejch has the highest estimated Bradley-Terry
strength, but further analysis with \code{equalp.test()} reveals that
there is no strong evidence in the datast to reject the hypothesis of
equal competitive strength ($p=0.26$), or that Vadlejch has a strength
higher than the null value of $\frac{1}{8}$ ($p=0.1$).

\section{Formula 1 motor racing: the constructors' championship}

In Formula 1 motor racing, the constructors' championship takes place
between {\em manufacturers} of racing cars (compare the drivers'
championship).  Here, we consider the constructor to be the object of
inference.  Each constructor typically fields two cars, each of which
separately accumulates ranking-based points at each venue.  Here we
use generalized Plackett-Luce to assess the contructors' performance.
The following table shows rankings for the first 9 venues of the 2021
season:

<<showconstructortable>>=
constructor_2021_table[,1:9]
@

Above, we see that Mercedes (``\code{Merc}'') came first and third at
Bahrain (\code{BHR}, and at Emilia Romagna (\code{EMI} came second and
retired (full details of the notation and conventions is given in the
package at \code{constructor.Rd}).  The identity of the driver is not
admissible information and indeed may change during a season.
Alternatively, we may regard the driver and the constructor as a joint
entity, with the constructor's ability to attract and retain a skilled
driver being part of the object of inference.

The associated generalized Plackett-Luce \code{hyper3} object is as follows:

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{constructor_2021}
\end{alltt}
\begin{verbatim}
## log( (AMM=1)^38 * (AMM=1, AR=1)^-1 * (AMM=1, AR=1, ARRF=1, ATH=1, HF=2,
## MM=1, Merc=1, WM=2)^-1 * (AMM=1, AR=1, ARRF=1, ATH=1, HF=2, MM=1,
## WM=2)^-2 * (AMM=1, AR=1, ARRF=1, ATH=1, HF=2, MM=2, Merc=1, WM=2)^-1 *
## (AMM=1, AR=1, ARRF=1, ATH=1, HF=2, RBRH=1, WM=1)^-1 * (AMM=1, AR=1,
## ARRF=1, ATH=1, HF=2, RBRH=1, WM=2)^-1 * (AMM=1, AR=1, ARRF=1, ATH=1,
## HF=2, WM=2)^-1 * (AMM=1, AR=1, ARRF=1, ATH=2, Ferrari=1, HF=2, WM=1)^-1
## * (AMM=1, AR=1, ARRF=1, HF=2, Merc=2, RBRH=1, WM=2)^-1 * (AMM=1, AR=1,
## ARRF=1, HF=2, WM=1)^-1 * (AMM=1, AR=1, ARRF=2, ATH=1, Ferrari=1, HF=2,
## MM=1, Merc=1, RBRH=1)^-1 * (AMM=1, AR=1, ARRF=2, ATH=1, Ferrari=1,
## HF=2, MM=1, Merc=1, RBRH=1, WM=1)^-1 * (AMM=1, AR=1, ARRF=2, ATH=1,
## Ferrari=1, HF=2, MM=1, Merc=2, RBRH=1, WM=2)^-1 * (AMM=1, AR=1, ARRF=2,
...
\end{verbatim}
\end{kframe}
\end{knitrout}
